---
output: md_document
title: "Spatial networks in R with sf and tidygraph"
---

The R community can lean on very powerful packages for spatial vector data analysis ([sf](https://github.com/r-spatial/sf)) and network analysis ([tidygraph](https://github.com/thomasp85/tidygraph)), that both support the popular 'tidy' approach to data science. In sf, spatial vector data are stored as objects of class `sf`, which are flat tables with a list column that contains the geometry of the features. Tidygraph is build on top of the widely-used [igraph](https://github.com/igraph/igraph) package, and stores networks in objects of class `tbl_graph`. A `tbl_graph` is an igraph-object, but enables the user to manipulate both the edges and nodes elements as being flat tables.

Despite the existence of sf and tidygraph, R seems to lack a general, modern way to store networks whose nodes are embedded in space, i.e. spatial networks. The [stplanr](https://github.com/ropensci/stplanr) package contains the `SpatialLinesNetwork` class. This, however, is based on [sp](https://github.com/edzer/sp/), a package for spatial data analysis launched in 2005, that is used less since sf entered the stage. The same yields for [spnetwork](https://github.com/edzer/spnetwork), a package that combined sp and igraph. More recently, [dodgr](https://github.com/ATFutures/dodgr) was created. This package provides very fast analytical tools for spatial networks, but deals solely with dual-weighted directed graphs, mainly used for calculating shortest paths in street networks.

This blogpost presents a general approach to store spatial networks in a tidy way by combining sf and tidygraph. At the same time, improvement points to make this approach more convenient, are discussed.

## FTW data example

Fietstelweek is, translated from Dutch, the National Bicycle Count week. It is a crowdsourced initiative to collect data about daily bike trips which will serve to improve the cycling infrastructure of the national, regional, and local bike network in the Netherlands.

The data was collected during a week, between September 14th and 20th, for 2015, and between September 19th and 25th for 2016. The data collection is performed through a mobile app which makes use of the cellphone GPS to track the trips.

Around 50 thousand people have participated in this joint effort gathering data on what routes cyclists take to get from their origins to their destinations, their speed, at what time, how many people use the same routes, and where the major delays are. 

The data is effectively available on [bikeprint.nl](http://www.bikeprint.nl/fietstelweek/) for download, and includes spatial files for edges and nodes (conforming a network), and a table for routes during different times of the day for the seven days.

The data contains what would be viewed as an ideal spatial network, including 'edges' and 'nodes', as well as 'route' data, that will not be discussed in this blog. To illustrate this, we can look into the municipality of Groningen. 

```{r}
library(piggyback)
pb_download("groningen_edges.gpkg")
pb_download("groningen_nodes.gpkg")
```

```{r}
library(sf)
edges <- st_read("groningen_edges.gpkg")
nodes <- st_read("groningen_nodes.gpkg")
```

The edge data is loaded as a `MULTILINESTRING` geometry, while the nodes are  `POINT` geometries. 

The `nodes` object contains a list of nodes with their codes (`KNOOPNUMME`), along with the average waiting time at each node (`TIJD`), and the number of routes that passed that node (`AANTAL`). 

```{r}
nodes
```

The `edges` contain a code for each edge (`LINKNUMMER`), the source and target nodes, the functional type of the edge (`HIGHWAY`), the average speed, and the number of routes using that edge (`INTENSITEI`).

```{r}
edges
```

