---
output: md_document
title: "Spatial networks in R with sf and tidygraph"
---

The R community can lean on very powerful packages for spatial vector data analysis ([sf](https://github.com/r-spatial/sf)) and network analysis ([tidygraph](https://github.com/thomasp85/tidygraph)), that both support the popular 'tidy' approach to data science. In sf, spatial vector data are stored as objects of class `sf`, which are flat tables with a list column that contains the geometry of the features. Tidygraph is build on top of the widely-used [igraph](https://github.com/igraph/igraph) package, and stores networks in objects of class `tbl_graph`. A `tbl_graph` is an igraph-object, but enables the user to manipulate both the edges and nodes elements as being flat tables.

Despite the existence of sf and tidygraph, R seems to lack a general, modern way to store networks whose nodes are embedded in space, i.e. spatial networks. The [stplanr](https://github.com/ropensci/stplanr) package contains the `SpatialLinesNetwork` class. This, however, is based on [sp](https://github.com/edzer/sp/), a package for spatial data analysis launched in 2005, that is used less since sf entered the stage. The same yields for [spnetwork](https://github.com/edzer/spnetwork), a package that combined sp and igraph. More recently, [dodgr](https://github.com/ATFutures/dodgr) was created. This package provides very fast analytical tools for spatial networks, but deals solely with dual-weighted directed graphs, mainly used for calculating shortest paths in street networks.

This blogpost presents a general approach to store spatial networks in a tidy way by combining sf and tidygraph. At the same time, improvement points to make this approach more convenient, are discussed.

## FTW data example

Fietstelweek is, translated from Dutch, the National Bicycle Count week. It is a crowdsourced initiative to collect data on what routes cyclists take to get from their origins to their destinations. The data were collected between September 14th and 20th in 2015, and between September 19th and 25th in 2016, through a mobile app which makes use of the cellphone GPS to track the trips. Around 50 thousand people participated. The data are available on [bikeprint.nl](http://www.bikeprint.nl/fietstelweek/) for download, and include seperate spatial files for edges and nodes.

To illustrate how such a spatial network could be stored in R, we use the Fietstelweek data for the municipality of Groningen as an example.

```{r}
library(piggyback)
pb_download("groningen_edges.gpkg")
pb_download("groningen_nodes.gpkg")
```

```{r}
library(sf)
edges <- st_read("groningen_edges.gpkg")
nodes <- st_read("groningen_nodes.gpkg")
```

Each node has a `POINT` geometry and several attributes: a unique ID (`KNOOPNUMME`), the average waiting time at the node (`TIJD`), and the number of routes that passed the node (`AANTAL`). 

```{r}
nodes
```

Each edge has a `MULTILINESTRING` geometry and several attributes: a unique ID (`LINKNUMMER`), the ID's of the source and target nodes, the functional type of the edge (`HIGHWAY`), the average speed, and the number of routes using that edge (`INTENSITEI`).

```{r}
edges
```
