---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# spnethack

The goal of spnethack is to explore spatial networks in R.

# Requirements

We'll use the following packages:

```{r, eval=TRUE, message=FALSE}
library(sf)
library(osmdata)
library(dodgr)
library(stplanr)
library(dplyr)
library(piggyback)
library(tidygraph)
```


# Data

It makes sense to have some input data.

## From josm

Data from OSM was downloaded with the `josm` GUI.
It can be read-in as follows:

```{r}
pb_download("promenade-all.geojson")
promenade_all = read_sf("promenade-all.geojson")
summary(factor(promenade_all$highway))
promenade_min = promenade_all %>% 
  filter(name == "Promenade")
promenade_way = promenade_all %>% 
  filter(!is.na(highway))
# write_sf(promenade_way, "promenade-way.geojson")
# write_sf(promenade_min, "promenade-min.geojson")
```

The minimum dataset can be read-in as follows:

```{r plot1, eval=TRUE}
pb_download("promenade-min.geojson")
promenade_min = read_sf("promenade-min.geojson")
plot(promenade_min$geometry)
```

A slightly larger dataset can be read-in and plotted as follows:

```{r pway, eval=TRUE}
pb_download("promenade-way.geojson")
promenade_way = geojsonsf::geojson_sf("promenade-way.geojson")
plot(promenade_way$geometry)
```

## From osmdata 

```{r}
promenade_osmdata = opq(bbox = 'Muenster, DE') %>% 
  add_osm_feature(key = 'name', value = 'Promenade') %>% 
  osmdata_sf %>% 
  unique_osmdata()
```

## From dodgr

```{r}
muenster = dodgr_streetnet('Muenster, DE')
promenade_dodgr = muenster %>% filter(name == 'Promenade')
```

## Route networks with stplanr

The code to create route network data is in `stplanr-promenade.R`.
It can generate a 'betweenness' graph like this:

```{r promenade-stplanr, eval=TRUE}
source(file = "stplanr-promenade.R")
```

## Route networks with dodgr

An example of how to create route data from sample points along the edges can be found `dodgr-promenade.R`, it can also generate a flow aggregate which is quite similar to the betweenness.

```{r promenade-dodgr, eval = TRUE}
source(file = "dodgr-promenade.R")

dodgr_flowmap(graph_f, linescale = 5)
```

To do this we can use all the points in the data 430 points on the `to` column and the same number on the `from`, and it takes 0.6 sec.


## Route networks with sfnetworks


## Route networks with spnetwork


## Route networks with osmnx


## Route networks with tidygraph

```{r tidygraph, eval=TRUE}
rtg = as_tbl_graph(x = rstp@g)
class(rtg)
length(rtg)
rtg$geometry = promenade_min$geometry
plot(rtg)
plot(rtg$geometry)

# But fails with subsetting...
rtg_sub = rtg[1:5, ]
rtg_sub
```

## Converting tidygraph into sf objects

```{r}
rtg_edges = activate(rtg,edges) %>% mutate(geometry = promenade_min$geometry)

st_as_sf(as.tibble(rtg_edges)) 
```

